------

## Notebook Instructions

------

* In many cases we have access to data with location information e.g. Address, City ,County ,Zip Code , Country. To visualize this data with more detail we need to use geolocation.

* We'll use Nominatim, a search engine for OpenStreetMap data, to easily find out geolocation coordinates.

* https://www.openstreetmap.org/about

### Load Packages in R/RStudio 

We are going to use tidyverse, a collection of R packages designed for Data Science. As well as other packages to help with data cleaning and processing. 

* tidyverse: https://www.tidyverse.org/
* rvest: https://github.com/tidyverse/rvest
* stringi: https://github.com/gagolews/stringi
* jsonlite: https://github.com/jeroen/jsonlite

```{r, echo = FALSE, message=FALSE, warning=FALSE}

options(scipen = 9999)

# Here we're checking if the package is installed
if(!require("tidyverse")){
  # If the package is not in the system then it will be install
  install.packages("tidyverse", dependencies = TRUE)
  # Here we are loading the package
  library("tidyverse")
}

# Here we're checking if the package is installed
if(!require("rvest")){
  # If the package is not in the system then it will be install
  install.packages("rvest", dependencies = TRUE)
  # Here we are loading the package
  library("rvest")
}

# Here we're checking if the package is installed
if(!require("stringi")){
  # If the package is not in the system then it will be install
  install.packages("stringi", dependencies = TRUE)
  # Here we are loading the package
  library("stringi")
}

# Here we're checking if the package is installed
if(!require("lubridate")){
  # If the package is not in the system then it will be install
  install.packages("lubridate", dependencies = TRUE)
  # Here we are loading the package
  library("lubridate")
}


# Here we're checking if the package is installed
if(!require("jsonlite")){
  # If the package is not in the system then it will be install
  install.packages("jsonlite", dependencies = TRUE)
  # Here we are loading the package
  library("jsonlite")
}
```

------

## Geocoding: OpenStreetMaps API Search Function

------

The geocode function below allows us to make requests to Nominatim, a search engine for OpenStreetMap data.

* Nominatim: https://nominatim.openstreetmap.org/
* Resources: https://nominatim.org/release-docs/develop/api/Search

```{r}
# INPUT LOCATION VARIABLES
# [Address], [City] , [State] , [Zipcode]

geocode <- function(record_id, address, city, state, zipcode){
  # NOMINATIM SEARCH API URL
  src_url <- "https://nominatim.openstreetmap.org/search?q="
  
  city <- str_replace_all(string = city, pattern = "\\s|,", replacement = "+")
  
  # CREATE A FULL ADDRESS
  addr <- paste(address, city, state, zipcode, sep = "%2C")
  
  # CREATE A SEARCH URL BASED ON NOMINATIM API TO RETURN GEOJSON
  requests <- paste0(src_url, addr, "&format=geojson")
  
  # ITERATE OVER THE URLS AND MAKE REQUEST TO THE SEARCH API
  for (i in 1:length(requests)) {
    
    # MAKE HTML REQUEST TO API AND TRANSFORME HTML RESPONSE TO JSON
    response <- read_html(requests[i]) %>%
      html_node("p") %>%
      html_text() %>%
      fromJSON()
    
    # FROM THE RESPONSE EXTRACT LATITUDE AND LONGITUDE COORDINATES
    lon <- response$features$geometry$coordinates[[1]][1]
    lat <- response$features$geometry$coordinates[[1]][2]
    
    # CREATE A COORDINATES DATAFRAME
    if(i == 1) {
      loc <- tibble(record_id = record_id[i], 
                    address = str_replace_all(addr[i], "%2C", ","),
                    latitude = lat, longitude = lon)
    }else{
      df <- tibble(record_id = record_id[i], 
                   address = str_replace_all(addr[i], "%2C", ","),
                   latitude = lat, longitude = lon)
      loc <- bind_rows(loc, df)
    }
  }
  return(loc)
}

```

--------

## Data Source: San Francisco Buildings Permit Data

--------

This data set pertains to all types of structural permits. Data includes details on application/permit numbers, job addresses, supervisorial districts, and the current status of the applications.

* Source: https://data.sfgov.org/Housing-and-Buildings/Building-Permits/i98e-djp9

```{r}

data <- read_csv("src/building_permits.csv")

```

### Data inspection

```{r}

head(data)

```



```{r}
tail(data)
```



```{r}

colnames(data)

```

### Change Columns Names

```{r}

colremove = c("First Construction Document Date",
              "Structural Notification",
              "Number of Existing Stories",
              "Number of Proposed Stories",
              "Voluntary Soft-Story Retrofit",
              "Fire Only Permit","Existing Units",
              "Proposed Units","Plansets",
              "TIDF Compliance","Existing Construction Type",
              "Proposed Construction Type","Site Permit",
              "Supervisor District","Current Police Districts",
              "Current Supervisor Districts",
              "Current Status Date", "Permit Creation Date",
              "Analysis Neighborhoods","Lot","Location",
              "SF Find Neighborhoods","Unit","Block", "Permit Type",
              "Unit Suffix","Street Number Suffix",
              "Existing Construction Type Description")

data <- data[colnames(data)[1:47]] %>% select(-all_of(colremove))

data <- rename(data, "Permit Type" = "Permit Type Definition")

data <- rename(data, "Neighborhoods Boundaries" = "Neighborhoods - Analysis Boundaries")

```


### Columns to lowercase

```{r}

colnames(data) <- str_replace_all(tolower(colnames(data)), " ","_")
```


### Dataset Dimensions (num rows x num columns)

```{r}
dim_desc(data)

```


### Select Current Status Complete

```{r}

bldg_df <- data %>% filter (current_status == "complete")

head(bldg_df)
```

### Dataset Dimensions

```{r}

dim_desc(bldg_df)

```


### Data Preparation and Cleaning

```{r}

bldg_df <- na.omit(bldg_df)

dim_desc(bldg_df)

```

### String Cleaning and Pattern Replacement

```{r}

bldg_df$city <- "San Francisco"
bldg_df$state <- "CA"
bldg_df$record_id <- as.character(bldg_df$record_id)

bldg_df$proposed_construction_type_description <-
  bldg_df$proposed_construction_type_description %>% 
  str_remove_all(pattern = "\\(|\\)") %>% 
  str_replace_all(" ", "_") %>% 
  as_factor()

bldg_df$street_name <- str_remove(string = bldg_df$street_name, pattern = "^0+")

bldg_df$street_name <- str_replace_all(string = bldg_df$street_name,
                                       pattern = "Bay Shore",
                                       replacement = "Bayshore")

bldg_df$street_suffix <- str_replace_all(string = bldg_df$street_suffix,
                                       pattern = "Bl",
                                       replacement = "Blvd")

bldg_df$street_suffix <- str_replace_all(string = bldg_df$street_suffix,
                                       pattern = "Tr",
                                       replacement = "Ter")

bldg_df$changed_use <- bldg_df$existing_use != bldg_df$proposed_use
bldg_df$estimated_revised_diff <- bldg_df$revised_cost - bldg_df$estimated_cost
```


### Date Types

```{r}
bldg_df$issued_date <- bldg_df$issued_date %>% mdy() %>% as_date()
bldg_df$issued_year <- bldg_df$issued_date %>% year()
bldg_df$issued_month <- bldg_df$issued_date %>% month(label = TRUE)
bldg_df$completed_date <- bldg_df$completed_date %>% mdy() %>% as_date()
bldg_df$completed_year <- bldg_df$completed_date %>% year()
bldg_df$completed_month <- bldg_df$completed_date %>% month(label = TRUE)

bldg_df$issued_completed_days <- bldg_df$completed_date - bldg_df$issued_date 
bldg_df$issued_completed_years <- bldg_df$completed_year - bldg_df$issued_year 

bldg_df$filed_date <- bldg_df$filed_date %>% mdy() %>% as_date()
bldg_df$filed_issued_days <- bldg_df$issued_date - bldg_df$filed_date

bldg_df$permit_expiration_date <- bldg_df$permit_expiration_date %>% mdy() %>% as_date()
bldg_df$issued_expiration_days <- bldg_df$permit_expiration_date - bldg_df$issued_date
```



### Project Cost and Zoning 

To look deeper into changes in zoning and type of constructions, we want to select only projects that changed the building use. Also we'll look at developments with significant capital investment, in particular greater than half a million dollars. 

```{r}

bldg_df <- bldg_df %>% arrange(desc(estimated_cost))
bldg_df <- bldg_df[bldg_df$changed_use == TRUE,]
bldg_df <- bldg_df %>% filter(revised_cost > 500000)

bldg_df$zipcode <- as_factor(bldg_df$zipcode)
bldg_df$permit_type <- as_factor(bldg_df$permit_type)
bldg_df$neighborhoods_boundaries <- as_factor(bldg_df$neighborhoods_boundaries)
bldg_df$proposed_use <- as_factor(bldg_df$proposed_use)

head(bldg_df)

```


### Final dataset dimensions


```{r}

dim_desc(bldg_df)

```


### Preparing the address string variable to match the OpenStreetMaps API

```{r}
# ADDRESS VARIABLE MUST MATCH NOMINATIM API
address <- paste0(bldg_df$street_number," ",
                  bldg_df$street_name," ",
                  bldg_df$street_suffix)

# DATA CLEANING SPECIAL CASES (e.g. 3333-3339 N CLARK)
query <- stri_replace(str = address, 
                      replacement = "", 
                      regex = "(-[0-9]+\\s)")

# REPLACE SPACES (\\s) OR COMMAS (,) WITH + SIGN
query <- str_replace_all(string = query, 
                         pattern = "\\s|,", 
                         replacement = "+")

head(query)
```


## Data Collection: OpenStreetMaps API Request 

```{r}
df <- geocode(record_id = bldg_df$record_id,
              address = query,
              city = bldg_df$city, 
              state = bldg_df$state,
              zipcode = bldg_df$zipcode)

head(df)
```

Create a main dataset with all the relevant information

```{r echo = FALSE, message=FALSE, warning=FALSE}

bldg_df <- left_join(bldg_df, df, by="record_id")

bldg_df$address <- str_replace_all(string = bldg_df$address,
                                   pattern = "\\+",
                                   replacement = " ")

colremove <- c("street_number","street_name", 
                "street_suffix", "current_status")

bldg_df <- bldg_df %>% select(-one_of(colremove))

head(bldg_df)

```


Finally let's save the new dataset containing the buildings geolocation (latitude, longitude).

```{r}
write_csv(bldg_df, "src/building_geocoord.csv")
```





